Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-05-19 overview.tex
Reynolds edits.

We begin this section by first writing down the complete set of
physical equations that be solved by \enzo, and then briefly describe
the numerical algorithms that we use to solve these equations.  This
section is intended to be an overview of the \enzo\ code's
capabilities: thus, we gather all of the equations solved into a
single place and provide a brief and high-level introduction to the
numerics of the code.  Detailed descriptions of the individual
components are then provided in Sections~\ref{sec.amr}-\ref{sec.num.analysis}.

@@

Any collisionless component (e.g. dark matter, stars) is modeled by particles which are governed by Newton's equations in comoving coordinates:
%
\begin{eqnarray}
\frac{d \vecx}{dt} 
    & = & \frac{1}{a} \vecv, 
          \label{eq:dm_position} \\
\frac{d \vecv}{dt} 
    & = & - \frac{\dot{a}}{a} \vecv
          - \frac{1}{a} \nabla \phi, 
          \label{eq:dm_velocity} 
\end{eqnarray}
%
The particles also make a contribution to the potential through their contribution to Poisson's equation.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself.  We do this whenever a level has caught up to the
coarser parent level and consists of entirely rebuilding the grids
on that level and at finer resolutions.  This is done by applying the grid refinement
criteria to the grids on that level, flagging zones that require
extra grids.  These criteria depend on the physical problem being
simulated.  We have implemented a number of options, including shock
and steep gradient detectors, but in the problem described here,
employ one based on the mass within a cell, imitating the Lagrangian
nature of the SPH algorithm.  Once a grid has a set of flagged cells,
we run a machine-vision based algorithm \citep{Berger91} to find edges
and determine a good placement of subgrids. 
%  These subgrids must not overlap one another, must cover all flagged
% cells and their neighbouring cells, and be above a preset efficiency
% threshold, where the efficiency of a cell is defined as the ratio of
% flagged cells to total cells. 
Once these new subgrids have been identified, the
solution from the next coarser grid is interpolated (see below) in
order to initialize the values on the new grids.  Finally, any overlap
between these new subgrids and the old ones is identified and the
solution within the regions of overlap is copied to the new subgrids.
The entire procedure just outlined is then repeated on the new grids
and in this way the entire hierarchy (from the original level examined
and continuing on to finer levels) is rebuilt.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with third order-accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
 (i) Construct monotonic parabolic interpolation of cell average data, for each fluid quantity.
% \item 
(ii) Compute interface states by averaging the parabola over the domain of dependence for each interface
% \item 
(iii) Use interface data to solve the Riemann problem.
 %\item 
 (iv) Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}

This PPM implementation does an excellent job of capturing strong
shocks in a few cells.  Multidimensional schemes are built up by
directional splitting and produce a method that is formally second
order-accurate in space and time and which explicitly conserves mass,
linear momentum, and energy.  A variety of reconstruction methods and
Riemann solvers have been implemented.

@@

\subsubsection{Godunov MUSCL (HD) with Dedner divergence cleaning (MHD)}
This solver was developed by Wang \& Abel to attack problems in
magnetic field amplification during the formation of galaxies
\citep{Wang:2009a} and to understand the role proto-stellar jets for
the theory of star formation \citep{Wang:2009b}. It combines the standard
approach of Godunov \citep{Godunov1959} for finite volume techniques
with the method of lines as described as e.g. by
\cite{leveque2002finite} and \cite{toro-1997}. In addition, however,
it implements the hyperbolic divergence cleaning algorithm of
\cite{2002JCoPh.175..645D}. It supports multiple approximate Riemann
solvers and non-ideal equations of states. Consequently,
these suite of solvers can be used for hydro and magneto-hydrodynamic
simulations. It is also this class of solvers as well as a version of
the PPM hydro solver which have been ported to Nvidia's CUDA framework
allowing \enzo to take advantage of modern graphics hardware \citep{Wang:2010}.

@@

The last hydrodynamics method we briefly describe is the
finite-difference algorithm used in the \zeus\ code, as given in more
detail in \citet{Stone92a}.  For this reason, we call this the \zeus\
method, although the code is entirely independent of the \zeus\ code,
and only the hydrodynamical algorithm of \zeus\ is implemented in
\enzo -- the MHD and RHD schemes are not.

@@

The current implementation of self-gravity in \enzo\ uses a Fast
Fourier Technique \citep{Hockney88} to solve Poisson's equation on the
coarsest (Level 0) grid.  The advantage of using this method is that
it is fast, accurate, and 
naturally allows both periodic and isolated boundary conditions for
the gravity, choices which are very common in astrophysics and
cosmology.  On subgrids, we interpolate the boundary conditions from
that grid's parent (either the level 0 grid or some other subgrid).
The Poisson equation is then solved using a multigrid technique.  In
\enzo, self-gravity is optional.  Aside from calculating the
gravitational potential from the baryon fields and particles existing
in the simulation, there are also a number of options for static gravitational fields.

@@

Collisionless matter (dark matter, stars, etc.) is modeled with
particles that interact with the baryons only via gravity.  These
particles are advanced in a single time step using a drift-kick-drift
algorithm to provide second-order accuracy even in the presence of
changing timesteps.  Since the particles follow the collapse of
structure, they are not adaptively refined.  Nor are there duplicate
sets of particles for each level; instead, each particle is associated
with the highest refined level available at its position in space, and is
moved as the hierarchy is rebuilt.  Thus, a particle has the same
timestep and feels the same gravitational force as a grid at that refinement
level.

@@

\enzo\ includes the capability of following up to
12 chemical species using a non-equilibrium solver.  The species can
be turned on in sets, with the simplest model including just H, H$^+$,
He, He$^+$, He$^++$, and $e^-$, and more complete models adding first
species important for gas-phase molecular hydrogen formation: $H^-$,
H$_2$ and H$_2^+$, and then $HD$ formation: $HD$, $D$, $D+$.  The
cooling and heating due to these species is also included (see the
next section).  The solution of the rate equations is carried out
using a time-backwards differencing which insures stability; to
maintain accuracy we sub-cycle the rate equations with a shorter
timestep such that the electron and neutral fractions do not change by
more than 10\%.

@@

Enzo may operate in a number of modes with regard to radiative cooling
and heating.  In the simplest mode, with the multi-species flag turned
off (i.e. set to zero), the cooling rate is computed from a simple
temperature-dependent cooling rate, taken from \citet{SW87}.  If
chemistry is turned on, then the code can include cooling from all
species of Hydrogen and Helium (including H$_2$) -- this primordial cooling is
also treated using a backwards-time differencing scheme with
sub-cycling to maintain accuracy.  Finally, we have also recently
added metal cooling based on a set of multi-dimensional lookup tables
computed with the \texttt{Cloudy} code \citep{1998PASP..110..761F} as
described in \citet{2008MNRAS.385.1443S} and
\citet{2011ApJ...731....6S}.  Finally, we note that the cooling and
heating is (almost always) treated in the optically-thin limit,
although the code can also follow radiative transfer in a limited set
of energy bins (see below).

@@

The chemical networks and heating rates described in the previous
section can be affected by external radiation fields, and the code
includes a number of pre-set radiation backgrounds that are uniform in
space but can very in time.  These are generally based on the
redshift-dependent rates given in \citet{1996ApJ...461...20H} and
\citet{2012ApJ...746..125H}, but can also include a uniform H$_2$
photo-dissociating background that is either constant in time or
varying as in \citet{WiseAbel05}.

@@

A second option for radiative transfer is a moment-based method that
adds an additional field tracking the radiant energy density.
This field is evolved using the flux-limited diffusion method, which
transitions smoothly between streaming (optically thin) and opaque
limits and is coupled to the hydrogen ionization network.  The
resulting set of linear equations is solved using the parallel HYPRE
framework.  Full details for the enzo implementation can be found in
\citet{ReynoldsHayesPaschosNorman2009}.

@@

Heat conduction, both isotropic and anisotropic, can be included using
an sub-cycled, operator-split method.  The heat fluxes are computed
with simple second-order accurate finite differences and stability is
ensured by restricting the time step, and using flux-limiters where
appropriate.

>>>>

We begin this section by first writing down the complete set of
physical equations that can be solved by \enzo, and then briefly describe
the numerical algorithms that we use to solve these equations.  This
section is intended to be an overview of the \enzo\ code's
capabilities: thus, we gather all of the equations solved into a
single place and provide a brief and high-level introduction to the
numerics of the code.  Detailed descriptions of the individual
components are then provided in Sections~\ref{sec.amr}-\ref{sec.num.analysis}.

@@

Any collisionless component (e.g. dark matter, stars) is modeled by particles which are governed by Newton's equations in comoving coordinates:
%
\begin{eqnarray}
\frac{d \vecx}{dt} 
    & = & \frac{1}{a} \vecv, 
          \label{eq:dm_position} \\
\frac{d \vecv}{dt} 
    & = & - \frac{\dot{a}}{a} \vecv
          - \frac{1}{a} \nabla \phi, 
          \label{eq:dm_velocity} 
\end{eqnarray}
%
The particles also make a contribution to the gravitational potential through their contribution to Poisson's equation.

@@

Since interesting regions on the grid may move, the hierarchy must
adapt itself.  We do this whenever a level has caught up to the
coarser parent level and consists of entirely rebuilding the grids
on that level and at finer resolutions.  This is done by applying the grid refinement
criteria to the grids on that level, flagging zones that require
extra grids.  These criteria depend on the physical problem being
simulated -- see Section~\ref{sec:refinement_criteria} for more details.  Once a grid has a set of flagged cells,
we run a machine-vision based algorithm \citep{Berger91} to find edges
and determine a good placement of subgrids. 
%  These subgrids must not overlap one another, must cover all flagged
% cells and their neighbouring cells, and be above a preset efficiency
% threshold, where the efficiency of a cell is defined as the ratio of
% flagged cells to total cells. 
Once these new subgrids have been identified, the
solution from the next coarser grid is interpolated (see below) in
order to initialize the values on the new grids.  Finally, any overlap
between these new subgrids and the old ones is identified and the
solution within the regions of overlap is copied to the new subgrids.
The entire procedure just outlined is then repeated on the new grids
and in this way the entire hierarchy (from the original level examined
and continuing on to finer levels) is rebuilt.

@@

We begin with the direct-Eulerian PPM implementation.  This is an
explicit, higher-order-accurate version of Godunov's method for ideal
gas dynamics with third order-accurate piecewise parabolic monotonic
interpolation and a nonlinear Riemann solver for shock capturing.  It
advances the hydrodynamic equations in the following steps:
%\begin{enumerate}
% \item 
 (i) Construct monotonic parabolic interpolation of cell average data, for each fluid quantity.
% \item 
(ii) Compute interface states by averaging the parabola over the domain of dependence for each interface.
% \item 
(iii) Use interface data to solve the Riemann problem.
 %\item 
 (iv) Difference the interface fluxes to update the cell average quantities.
%\end{enumerate}

This PPM implementation does an excellent job of capturing strong
shocks in a few cells.  Multidimensional schemes are built up by
directional splitting and produce a method that is formally second
order-accurate in space and time and which explicitly conserves mass,
linear momentum, and energy.  \red{(REF?)}  A variety of reconstruction methods and
Riemann solvers have been implemented.  \red{(These are never described!)}

@@

\subsubsection{Godunov MUSCL (HD) with Dedner divergence cleaning (MHD)}
This solver was developed by Wang \& Abel to attack problems in
magnetic field amplification during the formation of galaxies
\citep{Wang:2009a} and to understand the role of proto-stellar jets for
the theory of star formation \citep{Wang:2009b}. It combines the standard
approach of Godunov \citep{Godunov1959} for finite volume techniques
with the method of lines as described by
\cite{leveque2002finite} and \cite{toro-1997}. In addition, 
it implements the hyperbolic divergence cleaning algorithm of
\cite{2002JCoPh.175..645D}. It supports multiple approximate Riemann
solvers and non-ideal equations of states. Consequently,
these suite of solvers can be used for hydro and magneto-hydrodynamic
simulations. It is also this class of solvers as well as a version of
the PPM hydro solver which have been ported to Nvidia's CUDA framework
allowing \enzo\ to take advantage of modern graphics hardware \citep{Wang:2010}.

@@

The last hydrodynamics method we briefly describe is the
finite-difference algorithm used in the \zeus\ code, as given in more
detail in \citet{Stone92a}.  For this reason, we call this the \zeus\
method, although the code is entirely independent of the \zeus\ code,
and only the hydrodynamical algorithm of \zeus\ is implemented in
\enzo\ -- the MHD and RHD schemes are not.

@@

The current implementation of self-gravity in \enzo\ uses a Fast
Fourier Technique \citep{Hockney88} to solve Poisson's equation on the
coarsest (Level 0) grid on each time step.  The advantage of using this method is that
it is fast, accurate, and 
naturally allows both periodic and isolated boundary conditions for
the gravity, choices which are very common in astrophysics and
cosmology.  On subgrids, we interpolate the boundary conditions from
that grid's parent (either the level 0 grid or some other subgrid).
The Poisson equation is then solved on every timestep using a multigrid technique on one
subgrid at a time.  In
\enzo, self-gravity is optional.  Aside from calculating the
gravitational potential from the baryon fields and particles existing
in the simulation, there are also a number of options for static gravitational fields.

@@

Collisionless matter (dark matter, stars, etc.) is modeled with
particles that interact with the baryons only via gravity.  These
particles are advanced in a single time step using a drift-kick-drift
algorithm\citep{Hockney88} to provide second-order accuracy even in the presence of
changing timesteps.  Since the particles follow the collapse of
structure, they are not adaptively refined.  Nor are there duplicate
sets of particles for each level; instead, each particle is associated
with the highest refined level available at its position in space, and is
moved as the hierarchy is rebuilt.  Thus, a particle has the same
timestep and feels the same gravitational force as a grid at that refinement
level.

@@

\enzo\ includes the capability of following up to
12 chemical species using a non-equilibrium solver.  The species can
be turned on in sets, with the simplest model including just H, H$^+$, 
He, He$^+$, He$^++$, and $e^-$, and more complete models adding first
species important for gas-phase molecular hydrogen formation: $H^-$,
H$_2$ and H$_2^+$, and then $HD$ formation: $HD$, $D$, $D+$.  (We note
that a reduced version of the chemistry, used only with the
flux-limited diffusion radiation transport, includes only hydrogen
species.)  The
cooling and heating due to these species is also included (see the
next section).  The solution of the rate equations is carried out
using one Jacobi iteration of an implicit Euler time discretization to
ensure stability; to
ensure accuracy we sub-cycle the rate equations with a shorter
timestep such that the electron and neutral fractions do not change by
more than 10\%.

@@

Enzo may operate in a number of modes with regard to radiative cooling
and heating.  In the simplest mode, with the multi-species flag turned
off (i.e. set to zero), the cooling rate is computed from a simple
temperature-dependent cooling rate, taken from \citet{SW87}.  If
chemistry is turned on, then the code can include cooling from all
species of Hydrogen and Helium (including H$_2$) -- this primordial cooling is
also treated using 
one Jacobi iteration of an implicit Euler time discretization to maintain accuracy.  Finally, we have also recently
added metal cooling based on a set of multi-dimensional lookup tables
computed with the \texttt{Cloudy} code \citep{1998PASP..110..761F} as
described in \citet{2008MNRAS.385.1443S} and
\citet{2011ApJ...731....6S}.  Finally, we note that the cooling and
heating is (almost always) treated in the optically-thin limit,
although the code can also follow radiative transfer in a limited set
of energy bins (see below).

@@

The chemical networks and heating rates described in the previous
section can be affected by external radiation fields, and the code
includes a number of pre-set radiation backgrounds that are uniform in
space but can vary in time.  These are generally based on the
redshift-dependent rates given in \citet{1996ApJ...461...20H} and
\citet{2012ApJ...746..125H}, but can also include a uniform H$_2$
photo-dissociating background that is either constant in time or
varying as in \citet{WiseAbel05}.

@@

A second option for radiative transfer is a moment-based method that
adds an additional field tracking the radiation energy density.
This field is evolved using the flux-limited diffusion method, which
transitions smoothly between streaming (optically thin) and opaque
limits and is coupled to the hydrogen ionization network.  The
resulting set of linear equations is solved using the parallel HYPRE
framework.  Full details for the \enzo\ implementation can be found in
\citet{ReynoldsHayesPaschosNorman2009}.

@@

Heat conduction, both isotropic and anisotropic, can be included using
a sub-cycled, operator-split method.  The heat fluxes are computed
with simple second-order accurate finite differences and stability is
ensured by restricting the time step, and using flux-limiters where
appropriate.

### 2013-05-19 parallel.tex
Reynolds edits.

When \enzo\ is run on a distrubuted platform, some of the data is
replicated to all processors, while some remains on a single processor.
A description of the entire hierarchy (the position and size of each grid, as well
some other meta data) is stored on each processor.  This enables each
grid to know what grids are near it in the hierarchy, regardless of
where the data is.  The actual Baryon data (density, velocity, energy,
and any chemistry data) and particle data are stored on only one
processor.  Storing the entire hierarchy on all processor does create
some overhead, but in practice this is not an issue until one reaches
the extremely large scale.
  The code handles load balancing 
on a level-by-level basis such that the workload on each level is 
distributed as uniformly as possible across all processors.  Spatial locality of 
grids is not forced during message passing, for maximum flexibility (though not
necessarily maximum efficiency).  
The MPI message passing library\footnote{http://www-unix.mcs.anl.gov/mpi/}
 is used to transfer data between processors.

>>>>

When \enzo\ is run on a distrubuted platform, some of the data is
replicated to all processors, while some remains on a single processor.
A description of the entire hierarchy (the position and size of each grid, as well
some other meta data) is stored on each processor.  This enables each
grid to know what grids are near it in the hierarchy, regardless of
where the data is.  The actual Baryon data (density, velocity, energy,
and any chemistry data) and particle data are stored on only one
processor.  Storing the entire hierarchy on all processors does create
some overhead, but in practice this is not an issue until one reaches
extremely large scales.
  The code handles load balancing 
on a level-by-level basis such that the workload on each level is 
distributed as uniformly as possible across all processors.  Spatial locality of 
grids is not forced during message passing, for maximum flexibility (though not
necessarily maximum efficiency).  
The MPI message passing library
 is used to transfer data between processors.
